---
title: "STA 206 Homework 4"
author: "Kevin Su"
date: "10/28/2021"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# Problem 6

### Part A
```{r}
p6 <- cbind(c(-.97, 2.51, -0.19, 6.53, 1), c(-.63, .18, -.84, 1.6, 0.33), c(-.82, .49, .74, .58, -.31))
row.names(p6) = 1:5
colnames(p6) = c("Y", "X1", "X2")
p6

intercept0 <- rep(1, nrow(p6))

y6 <- p6[,1]
x6 <- cbind(intercept0, p6[,2:3])

y6
x6

xtx <- t(x6)%*%x6
xty <- t(x6)%*%y6
xtxinv <- solve(xtx)

xtx
xty
xtxinv
```

### Part B

```{r}

bvec <- xtxinv%*%xty

bvec

```
B0hat = 1.265271
B1hat = 2.679724
B2hat = 1.233270

### Part C

```{r}
library(Matrix)
H <- x6 %*% xtxinv %*% t(x6)
H

rankMatrix(H)

```

rank(H) = 3, and rank(I-H) = 2

### Part D

```{r}
traceH = 0
for (i in 1:nrow(H))
{
  traceH = traceH + H[i,i]
}
traceH
```
The trace of H is 3, equal to the rank of H.

### Part E
```{r}
model6 <- lm(y6~p6[,2]+p6[,3])

model6$fitted.values
model6$residuals
anova(model6)


```

The SSE is 0.911, and the MSE is 0.456. The degrees of freedom of SSE is 2.

### Part F

```{r}
x6nonadd <- cbind(x6, p6[,2]*p6[,3])
colnames(x6nonadd)[4] = "X1X2"

x6nonadd #design matrix

xFx <- t(x6nonadd) %*% x6nonadd

HF <- x6nonadd %*% solve(xFx) %*% t(x6nonadd)
HF

rankMatrix(HF)

```

rank(H) = 4 for part F
rank(I-H) = 1 for part F

The rank of H is 1 greater than the one from part C as there is one additional parameter

### Part G

```{r}
BG <- solve(xFx) %*% t(x6nonadd) %*% y6
BG

```

### Part H
```{r}
modelH <- lm(y6 ~ X1+X2+X1X2, data = as.data.frame(x6nonadd))
modelH$fitted.values
modelH$residuals
anova(modelH)

```
The SSE = 0.012
The MSE = 0.012
The degrees of freedom of SSE is 1.

### Part I
The second model appears to fit the data better as it has a higher adjusted R-squared value.

