---
title: "Homework 1 STA 206"
author: "Kevin Su"
date: "10/7/2021"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

```{r}
#Problem 7A
x <- seq(1,100)
```

```{r}
#Problem 7B
w <- 0.5 * x + 2
```

```{r}
#Problem 7C
normsamp <- rnorm(100, mean = 0, sd = 5)
sd(normsamp) 
mean(normsamp)
hist(normsamp)
```

Problem 7C: Symmetric distribution centered around zero. Sample Standard deviation of 4.55 and Mean of -0.2. 

```{r}
#Problem 7D
y <- w + normsamp
```

```{r}
#Problem 7E
plot(x,y = y)
#Problem 7F
coeffs <- lm(y~x)
abline(coeffs)
```

```{r}
#Problem 7G
coeffsres <- residuals(coeffs)
MSE <- sum(residuals(coeffs)^2)/length(residuals(coeffs))
plot(x, coeffsres)
MSE
```
Problem 7G: More residual values closer to 0 than farther away, no clear trend (as expected)


Problem 7H: Repeating C through G tends to yield similar results. 

```{r}
#Problem 7I
b1s = vector(mode = "numeric", length = 1000)
mses = vector(mode = "numeric", length = 1000)
b0s = vector(mode = "numeric", length = 1000)

for (i in 1:1000)
{
normsamp2 <- rnorm(100, mean = 0, sd = 5)
y2 <- w + normsamp2

coeffs2 <-lm(y2~x)
b0s[i] <- as.numeric(coeffs2$coefficients[1])
b1s[i] <- as.numeric(coeffs2$coefficients[2])
mses[i] <- sum(residuals(coeffs2)^2)/length(residuals(coeffs2))
}

hist(b0s, main = "Histogram of Intercepts (b0)")
hist(b1s, main = "Histogram of Slopes (b1)")
hist(mses, main = "Histogram of MSEs")

mean(b0s)
mean(b1s)
mean(mses)

var(b0s)
var(b1s)
var(mses)
```

Problem 7I: MSEs are centered around 25 (squared standard deviation), slopes centered around 0.5 as indicated in part B, and similarly intercepts centered near 2.