---
title: "STA206 Fall 2021: Quiz3 Question 2"
output:
  html_document: default
  word_document: default
---



**Instruction**:  <br>

* In this question, you will be asked to perform some  tasks in R <br>
* You should submit a .html or .pdf or .docx file 



In *Quiz3_data_Q2.Rdata*, you will find a data set called **data**  with 6 variables: *Y, V1, V2, V3, V4, V5*. The goal is to use **data** to build a model of *Y* using  *V1,...,V5* as (potential) predictors. Note that, you may need to transform the response variable *Y*.

* **(a). Load the data into the R workspace. **<br>

```{r}
load("C:/Users/kevin/Downloads/Quiz3_data_Q2.Rdata")
```

* **(b). Check variable type for each variable and convert $V1$ to a factor. **<br>

```{r}
sapply(data, class)
data[,2] <- as.factor(data[,2])
sapply(data, class)
```

* **(c). Fit the 1st-order model. Does this model suggest any transformation? **<br> 

```{r}
Model1 <- lm(Y~., data = data)
summary(Model1)

library(MASS)

boxcox(Model1)

pie(table(data$V1), main = "V1")

par(mfrow = c(2,3))
hist(data$Y)
hist(data$V2)
hist(data$V3)
hist(data$V4)
hist(data$V5)

```

*(Type your answer here) (c):* Large confidence interval for box-cox procedure including lambda = 1 does not suggest any transformation of Y is necessary. The values of V2-V5 seem fine.


* **(d) Build a model that satisfies the following requirements and briefly describe how you reach this model:  **<br>
  i. The number of regression coefficients in the model is no more than $10$;
  ii. Model assumptions hold (reasonably) well; 
  iii. The adjusted coefficient of multiple determination is at least $0.89$.


*(Note: even if you can not get a model that satisfies all the requirements, you may still get partial credits if your steps make sense. So try your best!)*

```{r}
# (d). (Type your code in the space below)
library("leaps")

subs <- regsubsets(Y~.^2, data = data, nbest = 10, nvmax = 10, method = "exhaustive")
subsummary <- summary(subs)

subrsq2 <- cbind(subsummary$which, subsummary$adjr2)
colnames(subrsq2) <- c(colnames(subsummary$which), "RSq-adj")
subrsq2

summary(lm(formula = Y ~ V1 + V2 + V3 + V5 + V1:V4 + V4:V5, data = data))

```

*(Type your answer here) (d):* 
I used the Regsubsets function to explore all interaction terms and return the top 10 hits with a max number of variables up to 10 through best subset selection. Despite finding quite a lot of models with adjusted R-sq above 0.89, I had great difficulty encoding some V1 interaction terms while excluding others, so I settled on the model outputted in the code which is Y ~ V1 + V2 + V3 + V5 + V1:V4 + V4:V5, with V1:V4 and V4:V5 being the only interaction terms. As such here there are 9 variables (not including intercept) and the Adjusted R-squared for this model is 0.8905. I considered using the stepAIC function to search but I felt it was not in the spirit of the quiz to search using a different criterion even though it would certainly have yielded a better, more clear result. 